---
title: "Missing Data HW5: Please refer to Section 2.f), 2.g), 3.e), 3.f), 4.d) and 4.e) for this Assignment"
author: "Karen Cao"
date: "3/10/2019"
output: pdf_document
---

```{r, include = FALSE}
# Load all required dependencies
library(dplyr)
library(plyr)
library(ggplot2)
library(mice)
library(mi)
library(lattice)

# Set the seed
set.seed(1234)
```

# 1. Data and Objective
The dataset is a panel data contains average total fertility rate ("TFR") for women in *India* from 1960 to 2016. The data also contains India's GDP values and recall lag with respect to each TFR value. There're 3 numeric and 1 categorical variables in this dataset and they are:

* TFR: average total fertility for women in India (a.k.a. the average number of children a women will give birth to over her entire fertility period).
* Year: year ranging from 1960 to 2016.
* GDP: India's GDP in billion dollars.
* Recall: length of the recall period between time of data collection and the date for the TFR values. This variable is numeric in the original dataset but has been recoded for the purpose of this analysis. Recall is categorized as 1) "Immediate" if the recall period is less than or equal to 1 year, 2) "Fast" if the recall period is less than or equal to 4 years and greater than 1 year, and 3) "Slow" if the recall period is greater than 4 years. 

The objective is to explore the relationship between the average total fertility rate and the other variables in the dataset using regression while considering TFR as the dependent variable and the rest as independent variables. 

```{r}
# Read and examine the data
dat = read.csv(file = "tfr_india.csv", sep = ",")[, 2:5]
head(dat)

# Create "TFR_missing" and "Recall_missing", which indiate missingness of "TFR" and "Recall", respectively
# 0 indicates missing value and 1 indicates otherwise
dat$TFR_missing = ifelse(is.na(dat$TFR), 0, 1)
dat$Recall_missing = ifelse(is.na(dat$Recall), 0, 1)
```

# 2. Missing Data Imputation
Variables "TFR" and "Recall" have missing values in the dataset. 

* Use mean and regression imputation to fill in missing values in "TFR" (see Part 2.a) and 2.c), respectively).
* Use mode imputation to fill in missing values in "Recall" (see Part 2 b)).
* Use regression imputation with noise to fill in missing values in "TFR" (see Part 2.d)) and "Recall.bi", which is a binary variable recoded from 3-level ordinal varaible "Recall" (see Part 2.e)).
* Use library(mi) and library(mice) with 5 chains and 50 iterations with the default variable types (see Part 2.f) and g))

```{r}
# Create a function that performs imputation for missing values
# The "method" argument in the function specifies the method used for missing value imputation
imp = function(x, method){
  missing = is.na(x)
  x.obs = x[!missing]
  imputed = x
  imputed[missing] = method(x.obs)
  return (imputed) # Output the imputed vector
}
```

## a) Mean Imputation - "TFR"
```{r}
dat$tfr.mean.imp = imp(dat$TFR, method = mean)

# Visualize imputed "TFR"
ggplot(dat, aes(Year, tfr.mean.imp)) +
  geom_point(aes(colour = factor(TFR_missing))) +
  scale_color_manual(labels = c("Yes", "No"), values = c("#edc649", "#1d9b91")) +
  labs(title = "Average Total Fertility Rate of India from 1960 to 2016", 
       y = "TFR", colour = "Imputed?")
```

## b) Mode Imputation - "Recall"
```{r}
# Function that returns mode of a categorical variable
mode = function(x) {
  ta = table(x)
  tam = max(ta)
  if (all(ta == tam))
    mod = NA
  else
    mod = names(ta)[ta == tam]
  return(mod)
}

dat$recall.mode.imp = imp(dat$Recall, method = mode)

# Visualize imputed "Recall"
ggplot(dat, aes(recall.mode.imp)) + 
  geom_bar(aes(fill = factor(Recall_missing))) +
  scale_fill_manual(labels = c("Yes", "No"), values = c("#edc649", "#1d9b91")) + 
  labs(title = "Bar Chart of Recall Period", x = "Length of Recall Period", y = "Frequency", fill = "Imputed?")
```

## c) Regression Imputation - "TFR"
```{r}
# First of all, build a model, "fit1", using ONLY completely observed variables "Year" and "GDP"
fit1 = lm(TFR ~ Year + GDP, data = dat)

# Impute missing "TFR" values based on fit1
pred = predict(fit1, newdata = ic(dat))

# Create a function that fills in missing TFR values using regression imputation
reg.imp = function(x, pred.value){
  missing = is.na(x)
  imputed = x
  imputed[missing] = pred.value
  return (imputed)
}

dat$tfr.reg.imp = reg.imp(dat$TFR, pred.value = pred)

# Visualize imputed "TFR"
ggplot(dat, aes(Year, tfr.reg.imp)) +
  geom_point(aes(colour = factor(TFR_missing))) +
  scale_color_manual(labels = c("Yes", "No"), values = c("#edc649", "#1d9b91")) +
  labs(title = "Average Total Fertility Rate of India from 1960 to 2016", 
       y = "TFR", colour = "Imputed?")
```

## d) Regression Imputation with Noise - "TFR"
```{r}
# Generate noise and add it to the previous imputation
noise = rnorm(n = length(pred), mean = 0, sd = summary(fit1)$sigma)
pred.noise = pred + noise
dat$tfr.reg.imp.noise = reg.imp(dat$TFR, pred.value = pred.noise)

# Visualize imputed "TFR" using regression imputation with noise
ggplot(dat, aes(Year, tfr.reg.imp.noise)) +
  geom_point(aes(colour = factor(TFR_missing))) +
  scale_color_manual(labels = c("Yes", "No"), values = c("#edc649", "#1d9b91")) +
  labs(title = "Average Total Fertility Rate of India from 1960 to 2016", 
       y = "TFR", colour = "Imputed?")
```

## e) Regression Imputation without Noise - "Recall.bi"
Since there's no binary variable in the original dataset, I re-coded "Recall", which is a ordinal variable with three levels: "Immediate", "Fast" and "Slow", into a binary varaible called "Recall.bi". "Recall.bi" has two levels, "Immediate" and "Not Immediate", where "Fast" and "Slow" were combined into one level -- "Not Immediate". (0 - Not Immediate; 1 - Immediate)
```{r}
# Recode "Recall" into "Recall.bi"
dat$Recall.bi = revalue(dat$Recall, replace = c("Fast" = "Not Immediate", "Slow" = "Not Immediate"), warn_missing = FALSE)

# Fit another model "fit2" and predict missing "Recall.bi" values using "fit2" WITH noise
fit2 = glm(Recall.bi ~ Year + GDP, data = dat[dat$Recall_missing == 1, ], family = "binomial")
recall.bi.reg.imp.noise = round(predict(fit2, newdata = dat[dat$Recall_missing == 0,], type = "response") + noise, 0)

# Save imputed "Recall.bi"
dat$recall.bi.reg.imp.noise = reg.imp(dat$Recall.bi, pred.value = ifelse(recall.bi.reg.imp.noise == 1, "Immediate", "Not Immediate"))

# Visualize imputed "Recall.bi"
ggplot(dat, aes(recall.bi.reg.imp.noise)) + 
  geom_bar(aes(fill = factor(Recall_missing))) +
  scale_fill_manual(labels = c("Yes", "No"), values = c("#edc649", "#1d9b91")) + 
  labs(title = "Bar Chart of Recall Period (Binary)", x = "Length of Recall Period", y = "Frequency", fill = "Imputed?")
```

## f) Missing Data Imputation using library(mi)
```{r, message = FALSE, warning = FALSE}
# Create the missing data frame object
dat.mdf = missing_data.frame(dat[, 1:4])
# show(dat.mdf)

# Impute until converged using the default variable types
mi.imp = mi(dat.mdf, n.iter = 50, n.chains = 5)
mi.conv = mi2BUGS(mi.imp)

# Plot diagnostics
plot(mi.imp)
```

## g) Missing Data Imputation using library(mice)
```{r, message = FALSE, warning = FALSE}
mice.imp = mice(dat[, 1:4], maxit = 50, m = 5)
plot(mice.imp)
```

# 3. Regression Analysis
Perform regression analysis using 

* a) Complete cases,
* b) Imputed data where missing values in "TFR" were filled using mean imputation and those in "Recall" were filled using mode imputation,
* c) Imputed data where missing values in "TFR" were filled using regression imputation, and
* d) Imputed data where missing values in "TFR" and "Recall.bi" were filled using regression imputation *with noise*
* e) Imputed data where missing values in "TFR" and "Recall" were filled using library("mi")
* f) Imputed data where missing values in "TFR" and "Recall" were filled using library("mice")

In all analyses, "TFR" is the dependent variable and "Year", "GDP" and "Recall" are independent variables. 

## a) using Complete Cases
```{r}
# the lm() function uses complete cases as default
summary(lm(TFR ~ Year + GDP + Recall, data = dat))
```

## b) using Imputed Data -- Mean and Mode Imputation
```{r}
summary(lm(tfr.mean.imp ~ Year + GDP + recall.mode.imp, data = dat))
```

## c) using Imputed Data - Regression Imputation
```{r}
summary(lm(tfr.reg.imp ~ Year + GDP + Recall, data = dat))
```

## d) using Imputed Data - Regression Imputation with Noise - "TFR" & "Recall.bi"
```{r}
summary(lm(tfr.reg.imp.noise ~ Year + GDP + recall.bi.reg.imp.noise, data = dat))
```

## e) using Imputed Data - library("mi") - "TFR" & "Recall"
```{r}
mi.analysis = mi::pool(TFR ~ Year + GDP + Recall, data = mi.imp)
summary(mi.analysis)
```
The equation could be written as:
$$TFR = 171.49 - 0.08Year + 0.16(Reacall == Slow)$$

## f) using Imputed Data - library("mice") - "TFR" & "Recall"
```{r}
mice.analysis = with(mice.imp, lm(TFR ~ Year + GDP + Recall))
summary(pool(mice.analysis)) # pool together the results

# Check fmi using the rule of thumb
mice::pool(mice.analysis)$pooled$fmi / 30
```
The equation could be written as:
$$TFR = 144.48 - 0.07Year + 0.02(Recall == Immediate) + 0.25(Reacall == Slow)$$

# 4. Discussion
## a) Complete Cases vs. Mean/Mode Imputation
The linear regression results using complete cases and imputed data are quite different. In the regression using complete cases, all predictors have significant relationship with TFR while when regress using imputed data, "Year" is the only signifiant predictor. Furthermore, the three predictors explain ~98% of the variance in the data under the complete cases approach but they only explain ~52% of the variance using imputed data. This implies that using complete cases may be a better way to perform linear regression instead of using data imputed from mean (for the numeric variable TFR) and mode imputation (for the categorical variable Recall).

## b) Complete Cases vs. Regression Imputation
The linear regression results using complete cases and imputed data with missing TFR values imputed using regression imputation are *exact the same* because variable "Recall" and "TFR" are perfectly correlated in the original dataset (That is to say, "TFR" is missing whenever "Recall" is missing). Since the function "lm()" in R uses complete cases as default, the linear regressions in Part 3.a) and 3.c) use the same subset of data and therefore have the exact same results. However, as we can see from the plot in Part 2.c), imputing TFR using regression imputation makes a lot more sense compared to doing that using mean imputation (see Part 2.a)).

```{r}
summary(lm(tfr.reg.imp ~ Year + GDP + recall.mode.imp, data = dat))
```
I re-did the analysis after filling missng "Recall" values using mode imputation so that it used the full dataset instead of just the complete cases. The result shows that all predictors are statistically significant and "Year", "GDP" and imputed recall (using mode imputation) explain ~99% variance of the data. This supports the earlier arugument that for imputing missing values in variable "TFR", regression imputation makes more sense than mean imputation.

## c) Complete Cases vs. Regression Imputation with Noise - "TFR" & "Recall.bi"
```{r}
# Re-fit the complete cases linear regression using "Recall.bi" instead of "Recall"
summary(lm(TFR ~ Year + GDP + Recall.bi, data = dat))
```
The linear regression results using complete cases and imputed data with noise are very similar. In both models, "Year" and "GDP" are statistically significant predictors. In addition, both models explain ~98% of the variance in the data. However, coefficients of the predictors in the second model have smaller standard errors, indicaiting the second model is more precise.

## d) Complete Cases vs. Imputation using library ("mi") - "TFR" & "Recall"
The linear regression results using complete cases and imputed data using library("mi") are very similar. In both models, all predictors have significant relationship with TFR and their coefficients didn't change much. However, standard errors of the predictors decrease a bit, indicating the second model may be more precise.

## e) Complete Cases vs. Imputation using library ("mice") - "TFR" & "Recall"
The linear regression results using complete cases and imputed data using library("mice") are kind of alike. In both models, all predictors have significant relationship with TFR. However, magnitudes of the coefficients of the predictors have changed as well as standard errors of the coefficients (some have increased and some have decrease). Therefore, we can't say which model is more precise.
